\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2021-12-09 submission/manuscript.Rmd
update manuscript with PDS feedback

---
csl: mbio.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
bibliography: references.bib
---

@@

*Clostridioides difficile* infection (CDI) has become more common and severe in the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and the presence of toxin genes have been associated with disease severity. However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied. These models are missing input from the gut environment. CDI is dependent on perturbations of the gut bacterial community in order to colonize the gut. The gut microbiota also impairs *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. However, it is unclear if the gut bacteria affect CDI disease severity. Here we demonstrate that gut bacterial communities contribute to disease severity variation. We derived diverse gut communities by colonizing germ-free mice with human fecal communities. The mice were then infected with a single *C. difficile* RT027 isolate which resulted in mortality and histopathologic differences. The severity grouped by the human fecal community they received. Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe disease. Bacterial groups associated with dietary fiber degradation, such as *Coprobacillus*, were associated with reduced disease severity. Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C. difficile* isolate.

@@

*Clostridioides difficile* infection (CDI) can be asymptomatic, or range in severity from mild diarrhea to severe outcomes of surgical interventions and even death. Models that predict severity and guide treatment decisions are based on clinical factors and *C. difficile* characteristics. The gut microbiome plays a role in potecting against CDI, though has not been investigated for its effect on CDI disease severity. We demonstrate that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities. The results reveal groups of bacteria associated with both severe and more mild *C. difficile* infections. Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C. difficile* was first identified as the cause of antibiotic-associated psuedomembranous colitis [@Kelly2008]. Disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death. Treatment for CDI is driven by the disease severity. The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count >15,000 cells/mm^3 and/or a serum creatinine >1.5 mg/dL, and complicated disease based on systolic blood pressure <90, ileus or toxic megacolon [@McDonald2018]. Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI. Thus, we have limited ability to prevent patients from developing severe CDI. Ideally, we could identify patients that are high-risk of severe disease use that to adjust their treatments to halt the disease before it worsens.

Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Firstly, there is no widely accepted definition of severe disease [@Ressler2021]. While the IDSA guideline is frequently used, many studies develop their own severity definitions. Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013]. However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021]. Thus, it may be necessary to incorporate information in addition to patient and clinical factors to accurately idenitfy patients at risk of severe CDI.

CDI disease severity is the result of damage to the host. Severity has been associated with characteristics of *C. difficile* ribotypes, such as 027, and host factors [@Bauer2011]. Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016]. The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016]. In addition to the pathogenic potential of *C. difficile*, the response by the host also affects the disease severity. IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011]. In addition to controlling the infection, the immune response can also cause intestinal injury. Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015]. While we have have elucidated these major drivers of host damage in CDI, these do not explain the observed variation in disease severity.

Missing from our CDI disease model are the effects of gut bacteria. *C. difficile* interacts with the intestinal community in many ways. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020]. Once in the gut, the indigenous bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [@Abbas2020]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019]. While the interaction between the gut bacteria and *C. difficile* is established, the effect the gut bacteria on CDI disease severity is unclear. Recent studies that have begun to uncover this relationship but have either used clinical data which does not permit control of the host or *C. difficile* variation [@Seekatz2016] or murine experiments which lack variation in the bacterial communities [@Dieterle2020]. Recently, Tomkovich et al. showed the variation in the bacterial communities between mice from different mouse colonies and vendors resulted in different clearance rates of *C. difficile* [@Tomkovich2020]. Therefore it is clear the gut bacteria modulate the activity and infection dynmaics of *C. difficile*, we hypothesized that gut bacteria also contribute to the variation in CDI disease severity. Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the CDI disease severity.

@@

***C. difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  Based on our previous observation that variation in the murine microbiota associated with different *C. difficile* colonization dynamics [@Tomkovich2020], we sought to explore how variation in murine microbial communities affects the outcome of *C. difficile* infection. To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a microbial community in the intestine of germ-free mice [@Nagpal2018]. We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage. The human fecal samples selected were chosen to provide diverse communtiy structures and based on community clusters [@Schubert2014]. The gut community was allowed to establish itself for two weeks post inoculation [@Gillilland2012]. We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A). The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to either their respective donor community or any mice who received a fecal homogenate from a different donor (Figure 1B). Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class abundances in their gut bacterial communities.

Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C. difficile* infection. A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C. difficile* colonization [@Chen2008; @Schubert2015]. However, we wanted to avoid modifying the communties with an antibiotic to maintain their unique microbial compositions and ecological relationships. Since these communties came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], or the communities became susceptible through the murine gut colonization process, we decided to test if *C. difficile* was able to infect these mice without any antibiotic perturbation. We hypothesized that *C. difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection colonizing the murine gut. After two weeks post incoluation with human fecal communities, the mice were then challenged with $10^{3}$ *C. difficile* ribotype 027 human isolate 431 spores. The mice were followed for 10 days post-challenge and their stool was collected and plated for *C. difficile* colony forming units (CFU) to detect infection level. Suprisingly, communities from all donors were colonized (Figure 2). Only two mice were able to resist *C. difficile* infection and both received their community from the same donor, Donor C. The remaining mice demonstrated susceptiblity to *C. difficile* infection without the need for any perturbation. Thus, *C. difficile* infected the mice without needing to perturb the established human-derived gut microbial communities, allowing them to maintain their unique community structures.

**Infection severity varies by initial community.** After challenging the mice with *C. difficile*, we investigated the outcome from the *C. difficile* infection and its relationship to the initial community. We followed the mice for 10 days post-challenge for *C. difficile* CFU, toxin production, and mortality. Seven mice, from Donors A, B and E, were not initally colonized on the day after *C. difficile* challenge but were infected ($>10^{6}$) by the end of the experiment. All mice that received their community from Donor J through O suffered from severe disease, which became highly colonized one day post-challenge and moribund within 3 days post-challenge. The remaining mice, except the Donor C mice which were not infected, maintained *C. difficile* infection through the end of the experiment (Figure 2). At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and cecal tissue was collected and scored for histopathologic signs of disease and fecal pellets were assayed for toxin activity (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A). Non-moribund mice from Donors D through I had comparable toxin activity as the moribund mice. Additionally, not all moribund mice had toxin activity detected in their stool. This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020]. Next we examined the cecal tissue for histopathologic damage. For moribund mice, we observed high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C. difficile* ribotype 027 [@Cowardin2016]. As observed with toxin activity, the moribund mice had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B). For Donor groups A through I, where all mice were non-moribund, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.

**Microbial community members explain variation in CDI severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated specific bacterial populations that could explain the variation in disease severity. To this aim, we interogated the day 0 bacterial community for their relationship to our CDI severity measurements using linear discriminant analysis (LDA) effect size (LEfSe) analysis. We split the mice into groups by severity level based on their moribundity and histopathologic score. We dichotomized the histopathologic scores into high and low bins by splitting on the median score. This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the mice that would develop only low intestinal injury. To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B). Many genera the were associated with toxin production as we would expect from their association with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score and no detected toxin. Overall, communities without genera that demonstrate a linear trend between relative abundance and disease severity can be better understood by their association with toxin activity. Lastly, we tested for correlations between the histopathologic score and the endpoint relative abundances of bacterial operational taxonomic units (OTUs) (Figure 4C). The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A). Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin. Our analysis have revealed bacterial taxa that are associated with three different measures of CDI severity in our model: moribundity, histopathologic score, and toxin activity.

We next wanted to determine whether collectively, bacterial community membership and relative abundance could be predictive of CDI severity. We trained random forest models with the day 0 bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score. Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). Collectively, members of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A). *C. difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria*. Next, we assessed the ability of the community to predict moribundity. Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). Bacterial taxa belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* hasd the greatest effect on model accuracy. Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). The only other class that was decreased in moribund mice was *Clostridia*. Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic score dichotomized as in previous analysis, AUROC = 0.99, Figure S2). No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.

**Disease severity consistent by donor community across isolates.** Since Carlson et al. showed clinical disease severity varied by *C. difficile* isolate characteristics [@Carlson2013; @Carlson2015], we next wanted to test if the effect of the community on CDI severity is conserved across *C. difficile* isolates. To facilitate this analysis, we selected a subset of donor fecal samples based on their associated mouse transplant communities that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund. Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3). We inoculated mice via oral gavage of the human fecal slurry of either Donor F, G, or M and allowed two weeks for the a community to establish. We then challenged the mice with a different *C. difficile* isolate and all mice were infected. The moribundity and histopathologic score outcomes from challenge with the second *C. difficile* human isolate were comparable to the outcomes of challenges with *C. difficile* ribotype 027 human isolate 431 by their donor group (Figure 6).

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate allowed us to identify the effect of microbiome variation on *C. difficile* infection disease severity. Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity. We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when a second *C. difficile* isolate is used. The disease outcome from the *C. difficile* chalenge with the new isolate matched the expected disease severities. Therefore, the effect of the community variation on disease severity *in vivo* is greater than the effect of *C. difficile* isolate phenotype variation [@Carlson2013; @Carlson2015]. Overall, these results have demonstrated that the bacterial community contributes to the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011; @McDonald2018; @Zar2007]. Those levels are being driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin A, toxin B, and binary toxin and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

>>>>

---
csl: mbio.csl
fontsize: 12pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

*Clostridioides difficile* infection (CDI) disease severity has increased over the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and toxin genes have been associated with disease severity. However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied. These models lack data describing the gut microbiota. Perturbations to the gut microbiota are necessary for *C. difficile* to colonize the gut. The gut microbiota is thought to impair *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. It has been unclear whether the gut microbiota affect CDI disease severity. However, here we demonstrate that gut bacterial communities can contribute to disease severity. We derived diverse gut communities by colonizing germ-free mice with different human fecal communities. The mice were then infected with a single *C. difficile* RT027 clinical isolate which resulted in moribundity and histopathologic differences. The variation in severity was associated with the human fecal community that the mice received. Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes. Bacterial groups associated with fiber degradation, such as *Coprobacillus*, were associated with less severe outcomes. Finally, we showed that, when infected with a different *C. difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community. These data have demonstrated that the gut bacteria can influence CDI disease severity and may improve clinical models.

@@

*Clostridioides difficile* infections (CDI) can be asymptomatic or range in severity from mild diarrhea to severe outcomes including surgical interventions and death. Models that predict severity and guide treatment decisions are based on clinical factors and *C. difficile* characteristics. Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity. We demonstrated that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease outcomes. These results revealed groups of bacteria associated with both severe and mild *C. difficile* infection outcomes. Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have increased in incidence and severity since *C. difficile* was first identified as the cause of antibiotic-associated pseudomembranous colitis [@Kelly2008]. CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death. Treatment for CDI is driven by the disease severity. The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018]. Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI. Thus, we have limited ability to prevent patients from developing severe CDI. Ideally, we could identify patients that are high-risk of severe disease and adjust their treatments to halt the disease before it worsens.

Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. There is no widely accepted definition of severe disease [@Ressler2021]. While the IDSA guideline is frequently used, many studies develop their own severity definitions. Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013]. However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021]. Thus, patient and clinical factors have not been sufficient to accurately identify patients at risk of severe CDI and may benefit from additional information involved in the development of CDI disease.

Severity has been associated with characteristics of *C. difficile* ribotypes, such as 027, and host factors [@Bauer2011]. Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016]. The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016]. In addition to the pathogenic potential of *C. difficile*, the response by the host also affects the disease severity. IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011]. In addition to controlling the infection, the immune response can also cause intestinal injury. Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015]. While we have have elucidated these major drivers of host damage in CDI, these are also the drivers of features of the poorly performing prediction models.

Missing from CDI severity prediction models are the effects of the indigenous gut bacteria. *C. difficile* interacts with the gut community in many ways. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020]. Once in the gut, the indigenous bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [@Abbas2020]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019]. While the relationship between the gut bacteria and *C. difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear. Recent studies have begun to uncover this relationship but have either used clinical data, which does not permit control of the host or *C. difficile* variation [@Seekatz2016] or murine experiments, which lack variation in the bacterial communities [@Dieterle2020]. Recently, Tomkovich et al. showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C. difficile* [@Tomkovich2020]. Therefore, it is clear the gut bacteria modulate the activity and infection dynmaics of *C. difficile*. We hypothesized that gut bacteria also contribute to the variation in CDI disease severity. Here, we colonized germ-free C57BL/6 mice with human fecal communities to test whether the variability in the gut bacteria can produce and explain variation in disease severity.

@@

***C. difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018]. We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage. The human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014]. The gut communities were allowed to establish themselves for two weeks post inoculation [@Gillilland2012]. We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A). The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to the mice who received a different human fecal donor community (Figure 1B). The communities were primarily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class relative abundances in their gut bacterial communities.

Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C. difficile* infection. A typical mouse model of CDI requires pre-treatment with antibiotics, such as clindamycin, to become susceptible to *C. difficile* colonization [@Chen2008; @Schubert2015]. However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships. Since these communities came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C. difficile* was able to infect these mice without any antibiotic perturbation. We hypothesized that *C. difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection when colonizing the murine gut. Mice were challenged with 10$^{3}$ *C. difficile* ribotype 027 human isolate 431 spores. The mice were followed for 10 days post-challenge and their stool was collected and plated for *C. difficile* colony forming units (CFU) to determine the extent of the infection. Suprisingly, mice from all donors were colonized (Figure 2). Only two mice were able to resist *C. difficile* infection and both received their community from the same donor, N3. By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C. difficile* infection without further modification of the gut community.

**Infection severity varies by initial community.** After we challenged the mice with *C. difficile*, we investigated the outcome from the infection and its relationship to the initial community. We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality. Seven mice, from Donors N1, N2 and N5, were not colonized on the day after *C. difficile* challenge but were infected (>10$^{6}$) by the end of the experiment. All mice that received their community from Donor M1 through M6 succumbed to the infection and became moribund within 3 days post-challenge. The remaining mice, except the uninfected Donor N3 mice, maintained *C. difficile* infection through the end of the experiment (Figure 2). At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal pellets were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A). Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice. Additionally, not all moribund mice had toxin activity detected in their stool. This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020]. Next we examined the cecal tissue for histopathologic damage. Moribund mice had high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C. difficile* ribotype 027 [@Cowardin2016]. As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B). The non-moribund mice, Donor groups N1 through N9, had a range in tissue damage from none detected to similar levels as the moribund mice, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donor groups but were largely consistent within each donor group.

**Microbial community members explain variation in CDI severity.** We next interrogated the bacterial communities at the time of *C. difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity. We split the mice into groups by severity level based on their moribundity and histopathologic score. We dichotomized the histopathologic scores into high and low groups by splitting on the median score. This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury. To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B). Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that associated with no toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score. Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C). The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A). Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin. This analysis has revealed bacterial genera that associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.

We next determined whether collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome. We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin production, moribundity, and day 10 post-challenge histopathologic summary score. For predicting if toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). *C. difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A). Next, we assessed the ability of the community to predict moribundity. Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). THe features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*. Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2). No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores. These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and toxin activity of CDI.

**Disease severity consistent by donor community across isolates.** Since Carlson et al. showed clinical disease severity varied by *C. difficile* isolate characteristics, such as germination rate [@Carlson2013; @Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C. difficile* isolates. We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish. We then challenged the mice with a different *C. difficile* isolate and all mice were infected. The moribundity and histopathologic score outcomes when challenged with a different *C. difficile* human isolate were comparable to the outcomes of challenges with *C. difficile* ribotype 027 human isolate 431 by their donor group (Figure 6). For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate, ribotype 027 isolate 431, demonstrated the variation in the gut bacteria affects the *C. difficile* infection disease severity. Our analysis revealed an association between the relative abundance of bacterial community members and disease severity. We used the relative abundance patterns of community members associated with disease outcomes to select donor communities to test whether the disease severity was conserved when challenged with a different *C. difficile* isolate. The disease outcomes from the challenge with the second *C. difficile* isolate matched the outcomes of the challenge with *C. difficile* ribotype 027 isolate 431. Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C. difficile* isolate variation [@Carlson2013; @Carlson2015]. Overall, these results have demonstrated that the bacterial community contributed to the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms to mild symptoms, such as diarrhea, to severe symptoms, such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007]. Those levels are driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevelance in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

The state of the gut bacterial community determines the ability of *C. difficile* to colonize, persist in the intestine. *C. difficile* is unable to colonize an unperturbed healthy gut community and is only able to become established after a perturbation [@Schubert2015]. Once colonized, the different communities lead to different metabolic responses and dynamics of the *C. difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021]. The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017]. The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019]. Members of the bacterial community also affect other metabolites *C. difficile* utilizes. *Bacteroides thetaiotaomicron* produce sialidases which release sialic acid from the mucosa for *C. difficile* to utilize [@Ng2013; @Ferreyra2014]. The nutrient environment affects toxin production [@MartinVerstraete2016]. Thus, many of the actions of the gut bacteria modulate *C. difficile* in ways that could affect the infection and resultant disease.
